### 2017-12-08 paper/sexSelectionSpacing-ver1.tex
Revise Data section
- Move substantial parts of recall discussion to online Appendix.
- Shorten spell definition part
- Reorder explanatory variables subsection
- General edits

% [Why India?]
I apply the method to birth histories for Hindu women, using data from 
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2006. 
India is a particularly interesting case.
On one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is clearly not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period 
\citep{Yi1993,park95}.
}
On the other hand, recent research suggests that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
}
In addition, there has been substantial changes in access and legality of
prenatal sex determination in India over the period covered.
Abortion has been legal in India since 1971, and still is, but the 
first reports of prenatal sex determination did
not appear until 1982--83 \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
The Act is described in detail at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach 5 convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Maharashtra was the first state to pass a similar law in 1988.
}
Hence, the data make it possible to show how spacing between births and
sex ratios have changed with the introduction of prenatal sex determination,
and whether the ban affected the relationship between birth spacing and sex 
ratios.

@@

I exclude visitors to the household, as well as
women married more than once, divorced, or not living with their husband,
women with inconsistent information on age of marriage,
and those with missing information on education.
Women interviewed in NFHS-3 who were never married or where gauna had not
been performed were also dropped.
The same goes for women who had at least one multiple birth,
reported having a birth before age 12, had a birth before marriage, or
a duration between births less than 9 months.
Women who reported less than 9 months between marriage and first birth
remain in the sample unless they are dropped for another reason.%
\footnote{
Women who report less than 9 months between marriage and first birth are retained 
because between 10 and 20 percent fall into this category.
Although it is possible that some of these births are premature the high number of
women who report a birth less than 6 months after marriage indicates that conception
likely occurred before marriage in most cases.
}

@@

Recall error arise mainly from child mortality, when respondents are reluctant to
discuss deceased children.%
\footnote{
The online Appendix contains a more thorough discussion of recall error and how I address it.
} 
Systematic recall error, where the likelihood of reporting a deceased child depends on
the sex of the child, is especially problematic because it biases the sex ratios.
Probing catches many missed births, but systematic recall error is still a potentially 
substantial problem.
Three factors contribute to the problem here.
First, girls have significantly higher mortality risk than boys.
Second, son preference may increase the probability that boys are remembered relative to girls.
Finally, in NFHS-1 and NFHS-2 enumerators probed only for a missed birth if the
initial reported birth interval was four calendar years or more.
But, given short durations between births, especially after the birth of a girl,
that procedure is unlikely to pick up all missed children.

Observed sex ratios by cohort provide a straightforward way to determine 
whether recall error is a problem.
Because prenatal sex determination techniques did not become widely available until the 
mid-1980s, a higher than natural sex ratio for cohorts born before that time must be 
the result of systematic recall error.
As shown in the online Appendix, the observed sex ratio by parity becomes more male 
dominated the further back births took place.
In addition, births in the same cohort tend to be more male dominated the more recent the 
survey (births in the cohort took place longer ago relative to the survey).
Hence, there is evidence of recall error and the degree of recall error increases
with length of time between survey and cohort.

Using cohort year of birth to analyze recall error and decide which observations
to keep is, however, problematic because the year of birth for a given parity is affected 
by recall error; for example, a second born child listed as first born will be 
born later than the real first born child.
Year of marriage should, however, be unaffected by recall error.
Using year of marriage the basic recall error pattern remains with women married longer 
ago more likely to report that their child was a boy for a given parity.
Similarly, comparing women married in the same period across surveys shows
that women married longer ago are more likely to report having sons.

That recall error increases the longer ago somebody was married means
that duration of marriage is a better predictor of recall error than calendar year of 
marriage.
Figure \ref{fig:sexRatioMarriage} shows the observed sex ratio for children 
reported as first born as a function of duration of marriage combining all three surveys.%
\footnote{
The graph for second births shows a similar pattern.
The graphs for the second births and the individual survey rounds are available upon request.
}
The solid line is the sex ratio of children reported 
as first born by the number of years between the survey and marriage, 
the dashed lines indicate the 95 percent confidence interval 
and the horizontal line the natural sex ratio (approximately 0.512).
To ensure sufficient cell sizes the years are grouped in twos.

% \begin{figure}[htp]%
% \centering
% \includegraphics[width=0.9\textwidth]{sex_ratio_marriage}
% \caption{Ratio of boys in ``first'' births}
% \label{fig:sexRatioMarriage}
% \end{figure}

Figure \ref{fig:sexRatioMarriage} clearly illustrates the systematic recall error
problem.
The observed sex ratio is increasingly above the expected value the
longer ago the parents were married.
The increasingly unequal sex ratio with increasing marriage duration suggests that
a solution to the recall error problem is to drop women who were married ``too far'' from 
the survey year.
The main problem is establishing what the best cut-off point should be.
The observed sex ratio is consistently significantly higher than the natural sex ratio 
from around 24 years of marriage, so one possibility is simply to drop all women married 
more than 24 years at the time of the survey.
But, as the Appendix shows, there are differences across the three surveys and between 
parities.
I therefore use different cut-off points by survey round.
For NFHS-1 women married 22 years or more were dropped, with the corresponding cut-off 
points 23 years for NFHS-2  and 26 years for NFHS-3.
The final sample consists of 146,096 women, with 332,951 parity one through four births.%
\footnote{
The online Appendix presents the results for a more restrictive definition and for the
women dropped because of recall error concerns.
}

@@

Spell duration is measured in months.
The first spell begins at the month of marriage because many 
women report giving birth less than 9 months after they were married.
For women who began living with their husbands at too young an age to conceive, the 
starting point should ideally be first ovulation, when she becomes ``at risk'' for a 
pregnancy, rather than month of marriage.
Unfortunately, information on age of menarche is only available in NFHS-1.
Instead, for women who began living with their husband before age 12, I set the 
the month they turned 12 years of age as staring point for the first spell.

The second and subsequent spells begin 9 months after the previous birth 
because that is the earliest we should expect to observe a new birth.
A few women report births that occurred less than 9 months 
after the previous birth; those women are dropped.

All spells continue until either a child is born or the spell is censored.
Censoring can happen for three reasons:
the survey takes place;
the woman is sterilized;
or the number of births observed becomes too sparse for the method to work.
The timing of censoring because of too few births vary slightly by spell but 
is generally 5 years or longer after the beginning of the spell.

I group spells into three time periods based on spell start date:
1972--1984, 1985--1994, and 1995--2006.
The first period covers the time before sex-selective abortions became widely available.
Abortion was legalized in 1971 and amniocentesis was introduced
in India in 1975, but the first newspaper reports on the availability of prenatal sex 
determination were not until 1982--83 \citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection became widespread
after a senior government official's wife aborted a fetus that turned out to be male \citep[p.\ 598]{Sudha1999}.
The second period covers the time from the widespread emergence of sex-selective abortions
until the prenatal Diagnostic Techniques (PNDT) act was passed in 1994.
The final period is from the PNDT act until the last available survey.
The PNDT act made it a criminal offence to reveal the sex of the fetus and was
followed by a campaign against the use of sex selection, although
enforcement appears to be relatively lax.

By dividing spells into these periods I can examine how the use of sex selection 
has changed across the three different regimes.
Note that the periods are based on the spells' beginning year, and some spells 
will therefore cover two periods.
A couple may, for example, be married in 1984, but not have their first child until 1986.
That couple's first spell will be in the 1972--1984 period, even though most of the 
spell actually falls in the 1985--1994 period.
Some children born from spells that began in the 1972--1984 period may therefore have been
conceive when prenatal sex determination techniques were available, which could result
in evidence of sex-selective abortions even for this period.
Similarly, a spell that began in the 1985--1994 period may have been partly or mostly
under the PDNT act.
The overall effect is to bias downwards any differences between the periods.

@@

The explanatory variables are divided into two groups.
The first group consists of variables expected to affect the shape of the hazard function: 
mother's education, sex composition of previous children, and area of residence.
Increasing the number of variables interacted with the baseline hazard lowers the risk 
of bias but requires more data to precisely estimate.
I chose these variables because the prior literature shows that they affect 
spacing choices and because the prior literature on sex selection indicates 
that these are correlated with sex selection.
The second group of variables are those expected to have an approximately 
proportional effect on the hazard: age of the mother at the beginning of the spell, 
length of her first spell (for second spell and above), whether the household owns 
land, and whether the household belongs to a scheduled tribe or caste.

Increasing education of mothers is strongly associated with lower fertility, with
the negative effect of higher opportunity cost on fertility more than outweighing the 
positive effect of higher income \citep{schultz97}.
Higher education should therefore be associated with higher use of sex selection.%
\footnote{
Fathers' education has two opposite predicted effects: the associated higher income
should increase fertility and therefore lower the pressure to use sex selection, but
the higher income also makes the use of sex selection cheaper.
In practice, fathers' education had little effect on the hazards and the use of 
sex-selective abortions and is not included.
}
I divide women into three education groups:
no education, 1 to 7 years of education, and 8 and more years of education.
The models are estimated separately for each education level.%
\footnote{
A potential concern here is reverse causation, where the sex of children
affect women's education.
Although no direct information is available on when women left school it is
possible to estimate school leaving age from the highest completed grade
and the usual starting age.
As described in the online Appendix, relatively few women could potentially 
have returned to school after the birth of their first child,
and only 56 could possibly have ended up in the wrong education group. 
Hence, there is little likelihood that reverse causation is a substantial
concern here.
}

@@

The sex composition of children, area of residence, and the interactions between
these are all interacted with the piece-wise linear baseline hazard dummies.
In other words, the baseline hazards are assumed to be different depending on
where a woman lives and the sex composition of her previous children.
As an example, for the second spell a separate regression is
run for each education level and in each regression four different baseline hazards 
are included (first child a boy in rural area, first child a boy in urban area,
first child a girl in rural area, first child a girl in urban area).
Although this approach substantially increases the number of regressions and 
estimated parameters it reduces the potential problem of including other variables 
as proportional effects.

The remaining variables are expected to affect hazards proportionally.
Although fecundity cannot be observed directly a suitable
proxy is the duration from marriage until first birth.
Most Indian women do not use contraception before the first birth
and there is pressure to show that a newly married woman can conceive 
\citep{dyson83,Sethuraman2007,Dommaraju2009}.
This is confirmed by the very short spells between marriage and first birth,
even among the most educated.
Hence, a long spell between marriage and first birth is likely due to low fecundity.
For both this variable and the age of the mother at the beginning of the spell 
the squares are also included.
The remaining variables are dummies for household ownership of land and membership
of a scheduled caste or tribe.

@@

Appendix Table \ref{tab:des_stat1} presents descriptive statistics for
the spells by education level and when the spell began.
There is a substantial number of censored observations.
As an example, for highly educated women who had their first child in the 1995--2006
period, almost half did not have their second child by the time of the survey.
Hence, although about 13,000 women began the second spell 
there are only about 7,000 births to these women.
Censoring becomes even more important for the third and fourth
spells, where around 70 percent of the observations are censored.
Generally, censoring increases with parity and time period.
This reflects a combination of factors: timing of the surveys
relative to the periods of interest, later beginning of childbearing, 
falling fertility, and the longer spells from sex-selective abortions.
The high number of censored observations underscore the importance of controlling for
censoring when examining the relationship between fertility and sex selection.

The descriptive statistics also provide a first indication of how the
sex ratio at birth changes over time and by spell.
For the first spell the sex ratio is very close to the natural
for all education groups and all three time periods.
As an example, among the highly educated group for the 1995--2006 period,
51.3 percent of the children born were boys.%
\footnote{
There still appears to be some recall error for the group of women without
education for the 1972--1984 period, where 52.3 percent of the children born were boys.
}
For the second spell, all but the highly educated group in the last two
periods have sex ratios in line with the natural sex ratio.
Women with 8 or more years of education have 53.1 and 54.3 percent
boys in the 1985--1994 and 1995--2006 periods, respectively.
This pattern repeats itself for the third spell, except the percentage
boys is higher for the high education group (55.3 and 55.9 for the last
two time periods).
Finally, for the fourth spell the high education group 
had 60 percent boys in the last period, i.e. after the PNDT act was introduced.
Note, however, that for the fourth spell the number of births is substantially
smaller and censoring even more important than for the other spells.

India's population has become progressively more urban.
For the first period, 32 percent of the women entering the first spell lived in urban areas.
This increases to 35 percent for the second period and to 42 percent for the final period.
The population is also substantially better educated.
Women with no education constituted almost 60 percent 
in the first period, but less than thirty percent in the last period.
Correspondingly, in the first period just over twenty percent had 8 or more 
years of education, but in the last period it was almost half.
Part of the increase in education is correlated with the increase in urbanization,
but the proportion of better-educated women has increased substantially
in the rural areas as well.
Among the high education group almost 70 percent lived in urban areas
during the first period but this had fallen to less than 60 percent
in the last period.

>>>>

% [Why India?]
I apply the method to birth histories for Hindu women, using data from 
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2006. 
India is a particularly interesting case.
On one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is clearly not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period 
\citep{Yi1993,park95}.
}
On the other hand, recent research suggests that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
}
In addition, there has been substantial changes in access and legality of
prenatal sex determination in India over the period covered.
Abortion has been legal in India since 1971, and still is, but the 
first reports of prenatal sex determination did
not appear until 1982--83 \citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection 
became widespread after a senior government official's wife aborted a 
fetus that turned out to be male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
The Act is described in detail at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach 5 convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Maharashtra was the first state to pass a similar law in 1988.
}
Hence, the data make it possible to show how spacing between births and
sex ratios have changed with the introduction of prenatal sex determination,
and whether the ban affected the relationship between birth spacing and sex 
ratios.

@@

I exclude visitors to the household, as well as
women married more than once, divorced, or not living with their husband,
women with inconsistent information on age of marriage,
and those with missing information on education.
Women interviewed in NFHS-3 who were never married or where gauna had not
been performed were also dropped.
The same goes for women who had at least one multiple birth,
reported having a birth before age 12, had a birth before marriage, or
a duration between births less than 9 months.
Women who reported less than 9 months between marriage and first birth
remain in the sample---unless they are dropped for another reason---because 
10 to 20 percent fall into this category across the 3 surveys.
Although it is possible that some of these births are premature, the high number of
women who report a birth less than 6 months after marriage indicates that conception
likely occurred before marriage in most cases.

@@

Recall error arise mainly from child mortality, when respondents are reluctant to
discuss deceased children.
Systematic recall error, where the likelihood of reporting a deceased child depends on
the sex of the child, is especially problematic because it biases both spacing
and sex ratios.
Probing catches many missed births, but systematic recall error is still a potentially 
substantial problem.
Three factors contribute to the problem here.
First, girls have significantly higher mortality risk than boys.
Second, son preference may increase the probability that boys are remembered relative to girls.
Finally, in NFHS-1 and NFHS-2 enumerators probed only for a missed birth if the
initial reported birth interval was four calendar years or more.
But, given short durations between births, especially after the birth of a girl,
that procedure is unlikely to pick up all missed children.
As I argue in the online Appendix, recall error is heavily dependent on the duration
between marriage and survey, and I therefore drop women married 22 years or more 
for NFHS-1, with the corresponding cut-off points 23 years for NFHS-2, 
and 26 years for NFHS-3.
The final sample consists of 146,096 women, with 332,951 parity one through four births.

% %
% \footnote{
% The online Appendix presents the results for a more restrictive definition and for the
% women dropped because of recall error concerns.
% }

@@

Spell duration is measured in 
As mentioned above, 10 to 20 percent of women gave birth less than 9 months
after they were married.
The first spell therefore begins at the month of marriage.
The exception is for women married very young, where I use the month they 
turned 12 rather than age of menarche, which is only available in NFHS-1.
The second and subsequent spells begin 9 months after the previous birth 
because that is the earliest we should expect to observe a new birth.
A few women report births that occurred less than 9 months 
after the previous birth; those women are dropped.

All spells continue until either a child is born or the spell is censored.
Censoring can happen for three reasons:
the survey takes place;
the woman is sterilized;
or too few births are observed for the method to work.
The timing of censoring because of too few births vary slightly by spell but 
is generally 5 years or longer after the beginning of the spell.

To examine how the use of sex selection has changed over time,
I group spells into three time periods based on spell start date:
1972--1984, 1985--1994, and 1995--2006.
These years follow the important changes in availability and legality of 
prenatal sex determination discussed in the Introduction.
Note that the periods are based on the spells' beginning year, and some spells 
will therefore cover two periods.
A couple may, for example, be married in 1984, but not have their first child until 1986.
That couple's first spell will be in the 1972--1984 period, even though most of the 
spell actually falls in the 1985--1994 period.
Some children born from spells that began in the 1972--1984 period may therefore have been
conceive when prenatal sex determination techniques were available, which could result
in evidence of sex-selective abortions even for this period.
Similarly, a spell that began in the 1985--1994 period may have been partly or mostly
under the PDNT act.
The effect is to bias downwards any differences between the periods.

@@

The explanatory variables are divided into two groups.
The first group consists of variables expected to affect the shape of the hazard function
(the $\mathbf{Z}$ variables): 
mother's education, sex composition of previous children, and area of residence.
I chose these variables because the prior literature shows that they affect 
spacing choices and are correlated with sex selection.
Increasing the number of variables interacted with the baseline hazard would
further lower the risk of bias but at the cost of requiring more data to 
precisely estimate.

Increasing education of mothers is strongly associated with lower fertility, with
the negative effect of higher opportunity cost on fertility more than outweighing the 
positive effect of higher income \citep{schultz97}.
Higher education should therefore be associated with higher use of sex selection.%
\footnote{
Fathers' education has two opposite predicted effects: the associated higher income
should increase fertility and therefore lower the pressure to use sex selection, but
the higher income also makes the use of sex selection cheaper.
In practice, fathers' education had little effect on the hazards and the use of 
sex-selective abortions and is not included.
}
I divide women into three education groups:
no education, 1 to 7 years of education, and 8 and more years of education.
The models are estimated separately for each education level to reduce
the potential problem of including other variables as proportional.%
\footnote{
A potential concern here is reverse causation, where the sex of children
affect women's education.
Although no direct information is available on when women left school it is
possible to estimate school leaving age from the highest completed grade
and the usual starting age.
As described in the online Appendix, relatively few women could potentially 
have returned to school after the birth of their first child,
and only 56 could possibly have ended up in the wrong education group. 
Hence, there is little likelihood that reverse causation is a substantial
concern here.
}

@@

The second group of variables are those expected to have an approximately 
proportional effect on the hazard: 
length of her first spell (for second spell and above to capture fecundity), 
age of the mother at the beginning of the spell, 
whether the household owns land, 
and whether the household belongs to a scheduled tribe or caste.
Fecundity cannot be observed directly, but a suitable proxy is the 
duration from marriage until first birth.
Most Indian women do not use contraception before the first birth
and there is pressure to show that a newly married woman can conceive,
as confirmed below by the very short spells between marriage and first birth,
even among the most educated \citep{dyson83,Sethuraman2007,Dommaraju2009}.
Hence, a long spell between marriage and first birth is likely due to low fecundity.
For this variable and the age of the mother at the beginning of the spell 
the squares are also included.
The remaining variables are dummies for household ownership of land and membership
of a scheduled caste or tribe.

@@

Appendix Table \ref{tab:des_stat1} presents descriptive statistics for
the spells by education level and when the spell began.
There is a substantial number of censored observations.
As an example, for highly educated women who had their first child in the 1995--2006
period, almost half did not have their second child by the time of the survey.
Hence, although about 13,000 women began the second spell 
there are only about 7,000 births to these women.
Censoring becomes even more important for the third and fourth
spells, where around 70 percent of the observations are censored.
Generally, censoring increases with parity and time period.
This reflects a combination of factors: timing of the surveys
relative to the periods of interest, later beginning of childbearing, 
falling fertility, and the longer spells from sex-selective abortions.

The descriptive statistics provide a first indication of how the
sex ratio at birth changes over time and by spell.
For the first spell the sex ratio is very close to the natural
for all education groups and all three time periods.%
\footnote{
There still appears to be some recall error for the group of women without
education for the 1972--1984 period, where 52.3 percent of the children born were boys.
}
For the second spell, all but the highly educated group in the last two
periods have sex ratios in line with the natural sex ratio.
Women with 8 or more years of education have 53.1 and 54.3 percent
boys in the 1985--1994 and 1995--2006 periods, respectively.
This pattern repeats itself for the third spell, except the percentage
boys is higher for the high education group (55.3 and 55.9 for the last
two time periods).
Finally, for the fourth spell the high education group 
had 60 percent boys in the last period, i.e. after the PNDT act was introduced.
Note, however, that for the fourth spell the number of births is substantially
smaller and censoring even more important than for the other spells.

India's population has become progressively more urban.
For the first period, 32 percent of the women entering the first spell lived in urban areas.
This increases to 35 percent for the second period and to 42 percent for the final period.
The population is also substantially better educated.
Women with no education constituted almost 60 percent 
in the first period, but less than thirty percent in the last period.
Correspondingly, in the first period just over twenty percent had 8 or more 
years of education, whereas in the last period it was almost half.
Part of the increase in education is correlated with the increase in urbanization,
but the proportion of better-educated women has increased substantially
in the rural areas as well.
Among the high education group almost 70 percent lived in urban areas
during the first period but this had fallen to less than 60 percent
in the last period.
